Towards an Ecological Trait-data Standard

### 2018-05-02 manuscript/manuscript.Rmd
update manuscript, argument for use, author list, bibliography

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: Markus Fischer \footnotemark[4]\footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Birgitta König-Ries\footnotemark[5]
- name: Nadja K. Simons \footnotemark[6]
csl: amnat.csl
fontsize: 10pt
geometry: margin=1in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F), Senckenberganlage 25, 60325 Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[6]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07]. Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity [@diaz16] and ecosystem functioning (refs). 
For example, using traits is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.

@@

In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13;  @ricklefs17].

Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org. 
Furthermore, these services warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12]. <!-- AG: Perhaps mention the Bouchout declaration here: http://www.bouchoutdeclaration.org/ ? -->
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.

Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management. 
As a result of this, datasets also greatly vary in structure. 
The column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives. Datasets vary in the detail provided in the metadata, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
If factual trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated quantitative measurements averaged from raw data, they usually are reported in a species$\times$trait wide-table format. Here, each row contains a species (or taxon) for which multiple traits are reported in columns. 
Similarly, when reporting raw data, researchers place observations of individual specimens in rows with multiple trait measurements applied to the same specimen placed in columns. 
The variability of the number and meaning of columns in these tables requires manual adjustments when merging multiple datasets.

@@

<!-- database backend, selective download, user access management --> 
More specialised trait-database platforms have been created that cover certain trait types (e.g. floral traits, seed traits or wood density traits), interaction types (e.g. pollination traits or feeding relationships), or environmental and experimental context of the trait observation (e.g. location or climatic data).  
These centralized database initiatives attract data submissions from a defined research field and take care of the harmonization process and greatly facilitate data synthesis. 
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
By aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other platform. The online platform enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
On the side of animals, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/). 
Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking. The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].

<!--Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12]. Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. -->

Regarding open access, hardly any of these centralised databases for traits comply with the criteria demanded by journals and funding agencies for primary data publication. They incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.  
Time will show how these incentives fare with an increasing availability of open access trait data  on low-threshold data repositories. 
<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

@@

In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives]. For example, the Common Daisy (*Bellis perennis*) would be related via a globally unique taxonID using the URI https://www.gbif.org/species/3117424 from the GBIF Backbone Terminology. The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review. 
Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/]. 
The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11].

@@

By filling this gap, data brokering services (e.g. the German Federation on for Biological Data; gfbio.org) or data management systems for scientific projects (e.g. BExIS, NEON<!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses. <!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->

@@

Term | Definition 
----- | -----
Terminology | 
Thesaurus | 
Ontology | 
semantic web | 
relational database | 
dataset | 
Web service | 
Online Portal |
Repository | 
Darwin Core Terms |
Darwin Core Archive |

@@

@@

As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees of measurement detail for multiple use cases of trait-based ecological research. 
Rather than describing a data structure for relational databases, this is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in a semantic web.

@@

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We paid particular consideration to the work of @kattge11, @kattge11a,  and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. 
However, the vocabulary proposed is meant as a preliminary draft of a standard nomenclature to be completed and corrected by the wider community of researchers using trait-based approaches in ecology, and may in the future integrate more morphological and evolutionary perspectives as well.
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

@@

- Offering compatibility, input for dimensional databases by using global identifiers and internal identifiers that provide hierarchical layers (measurement, occurrence, sampling event, data source, bibliographic reference)
- using these identifiers enables the vocabulary for for observation level data (linked via an occurrence identifier) raw measurement data (each row an own measurement identifier), as well as for aggregated data on the species level (each row an own measurement identifier plus aggregation information, like taxon rank or statistical method). allows for originary data (reporting own measuremetns) or compiled data sheets (reporting multiple dataset identifiers).  
- wherever appropriate, use URIs for linking to external ressources. 
<!-- PM: It seems that the object identifier for an individual is the key distinction here- this could be clearer.
Maybe you address this later but does your tool assign one of these- how to you make sure they are all unique if there are multiple users- e.g. if you pooled together two previously complied datasets which might contain some of the same individuals?
-->

- long-table format. has advantage of easy merging and filtering, computing, modelling. But disadvantage of multi-type columns (e.g. for values) and repetitive content.

- linking to standard taxonomy and trait terminology. semantic web standards, URI identifiers of external resources.

- Allowing for but minimizing duplicate information.  by providing standardised duplicates, align user-provided nomenclature and trait methodology with taxonomic nomenclature and trait ontologies.  transparency for data user.

- vocabulary is structured into sections: core, measurement, occurrence, metadata, traitlists. Short definitions. These sections are meant to guide through the vocabulary. their structural meaning depends on context provided by hierarchical IDs.

Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain. For datasets containing multi-layered information on observations, traits, taxa and environmental context, we highly recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr15, ideally compiled into Darwin Core Archives [DwC-A, @robertson09]. 
<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->

@@

In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider et al. 2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/". In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.

Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam14; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e. via the Application Programming Interface, API). Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

<!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear. It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc. -->
 <!-- Nadja: I like the idea of having a figure explaining the different terms -->

<!-- 
(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of the Entity-Quality Model. However, such network data may be of interest to be combined with trait data. ) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names. -->

@@

# Specification of traits

@@

Many terms of these extensions refine or copy terms of the Darwin Core Standard and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: Markus Fischer \footnotemark[4]\footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Anton Güntsch \footnotemark[5]
- name: Pete Manning \footnotemark[2]
- name: Birgitta König-Ries\footnotemark[6]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 12pt
geometry: margin=1in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F), Senckenberganlage 25, 60325 Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Botanischer Garten und Botanisches Museum Berlin-Dahlem, Freie Universität Berlin, Berlin, Germany}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity [@diaz16] and ecosystem functioning (refs). 
For example, using traits is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.

@@

In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15;  @renner14; @vanderplas17; @price13;  @ricklefs17].

Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net, Zenodo.org. 
Furthermore, these services warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, which supports the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.

Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management. 
As a result of this, datasets also greatly vary in structure. 
Data management initiatives that arise from scientific communities and projects, like the KNB (the Knowledge Network for Biocomplexity; https://knb.ecoinformatics.org/), offer means to meet the high scientific criteria of data management.

For trait-data, there are specific issues arising from the variability of data structure. 
For instance, the column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives. Datasets vary in the detail provided in the metadata, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
If factual trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated quantitative measurements averaged from raw data, they usually are reported in a species$\times$trait wide-table format. Here, each row contains a species (or taxon) for which multiple traits are reported in columns. 
Similarly, when reporting raw data, researchers place observations of individual specimens in rows with multiple trait measurements applied to the same specimen placed in columns. 
The variability of the number and meaning of columns in these tables requires manual adjustments when merging multiple datasets.

@@

<!-- database backend, selective download, user access management --> 
More specialised trait-database platforms have been created that cover certain trait types (e.g. floral traits, seed traits or wood density traits), interaction types (e.g. pollination traits or feeding relationships), or environmental and experimental context of the trait observation (e.g. location or climatic data).  
These centralized database initiatives attract data submissions from a defined research field and take care of the harmonization process and greatly facilitate data synthesis. 
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
By aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other platform. The online platform enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12]. Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. 
On the side of animals, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/). 
Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking. The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans, make harmonizing animal trait data a more complex task than for plant traits [@moretti17].

Regarding open access, hardly any of these centralised databases for traits comply with the criteria demanded by journals and funding agencies for primary data publication. The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.  
With the pro-active turn towards open access data (as declared inthe Bouchot Declaration; http://www.bouchoutdeclaration.org/) it may be necessary to find other incentives for data submission.

<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

@@

In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives]. The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review. 
Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/]. 
The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11].

@@

By filling this gap, data brokering services [e.g. the German Federation on for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g. BExIS, NEON Dataportal; Diversity Workbench, diversityworkbench.net; KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses. <!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy) -->

## Conclusion of review

Initiatives for standardization (e.g. ontologies and data standards) and platforms for data management (e.g. database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation.  
With low-threshold offers for the hosting of scientific primary data attracting a wealth of heterogeneous trait-datasets, the concern is with data harmonization of these distributed data sources.

The goal must be to integrate these distributed data better into the global biodiversity data-management ecosystem by creating awareness for data standardization on the side of data providers. 
We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in data visibility and interconnectedness of the published data.

@@

@@

Term | Definition 
----- | -----
Terminology |  
Term |
Thesaurus | 
Ontology | 
semantic web | 
relational database | 
two-dimensional data | 
dataset | 
Web service | 
Online Portal |
Repository | 
Darwin Core Standard |
Darwin Core Archive |
Application Programming Interface (API) | 
Data standard | 
Uniform Resource Identifier (URI) |

@@

As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees of measurement detail for multiple use cases of trait-based ecological research. 
Rather than describing a data structure for relational databases, this is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the 'semantic web'.

@@

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We paid particular consideration to the work of @kattge11, @kattge11a,  and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases.

However, the vocabulary proposed is meant as a preliminary draft of a standard nomenclature to be completed and corrected by the wider community of researchers using trait-based approaches in ecology, and may in the future integrate more morphological and evolutionary perspectives as well.
Also, at present we conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of the Entity-Quality Model. However, such network data may easily be combined with trait data via the taxon identifiers or by using other Extensions of the Darwin Core (DwC).
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

@@

The incentives for using a unified vocabulary for trait data might not be obvious for data providers, who are mainly concerned with a particular study site, organism group and research question. We won't discuss here in depth how sharing trait data will improve science on broader questions and larger scales, by providing a data basis for downstream analysis. It appears to be broad consensus that an open biodiversity science is crucial for an evidence based decision making and conservation policy on regional and global scales.

Instead, this section points on the technical aspects and advantages of readying own trait data to link with the 'semantic web' of biodiversity data and knowledge. 
By using a constrained vocabulary with globally accessible definitions of terms, data can be accessed more easily by other researchers and datasets from different sources can be combined into aggregated datasets.
Also it will be easier to develop computational methods and software tools that access and handle the data, based on the standard vocabulary.

It is implicit to this vocabulary to use a two-dimensional observation long-table format for the data, rather than a species-trait matrix (where some of the columns would be trait names which are not included in the vocabulary, but are part of trait thesauri). As this format draws from a defined set of columns,  merging of datasets is  easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data; @wickham14].

Using well-defined identifiers ('IDs') to structure the dataset offers compatibility with the input formats for larger databases that may offer better visibility and impact of the data contribution. 
Identifier fields are key elements to structure the two-dimensional dataset. 
For instance, for occurrence level data, were multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link these measurements across the rows of the dataset. 
Similarly, multivariate measurements, like for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
If literature data are provided, data are probably not structured below the taxon level and a 'taxonID' is the key identifier.
A 'datasetID' allows to report datasets that have been compiled from multiple sources.  
Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external ontology services (see above for resources).

However, two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain. For datasets containing multi-layered information on observations, traits, taxa and environmental context, we  recommend to consider relational datatabase structures composed of multiple spreadsheets linked via identifier columns, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a]. The trade-off is data-consistency and readability in a single table *vs.* avoidance of content duplication and redundancy in a relational database. The standard vocabulary proposed here may still be applied within these structured databases.

@@

In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider, F.D., Jochum, M.,  LeProvost, G., Penone, C., Ostrowski, A. and Simons, N.K., 2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/". In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.

Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam14; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e. via the Application Programming Interface, API **in preparation!**). Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

@@

### Specification of traits

@@

Many terms of these extensions refine or copy terms of the Darwin Core Standard (Dwc) and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
